---
title: "COVID-19 US Data Analysis"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(httr)  # For downloading files
```

## Data Loading

```{r load-data}
# URLs for US COVID-19 data
cases_us_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv"
deaths_us_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv"

# Download files if they don't exist locally
if (!file.exists("cases_us.csv")) {
  GET(cases_us_url, write_disk("cases_us.csv", overwrite = TRUE))
}
if (!file.exists("deaths_us.csv")) {
  GET(deaths_us_url, write_disk("deaths_us.csv", overwrite = TRUE))
}

# Read the downloaded CSV files
cases_us <- read.csv("cases_us.csv", check.names = FALSE)
deaths_us <- read.csv("deaths_us.csv", check.names = FALSE)

# Keep only relevant columns
cases_us <- cases_us %>% select(-c(iso2, iso3, code3, FIPS, Admin2, Lat, Long_, Combined_Key))
deaths_us <- deaths_us %>% select(-c(iso2, iso3, code3, FIPS, Admin2, Lat, Long_, Combined_Key))
```

## Data Processing

```{r data-processing}
# Convert wide format to long format while filtering last 6 months
latest_date <- max(names(cases_us)[-(1:11)], na.rm = TRUE)
latest_date <- mdy(latest_date)

cases_us_long <- cases_us %>%
  pivot_longer(cols = -(1:11), names_to = "date", values_to = "cases") %>%
  mutate(date = mdy(date)) %>%
  filter(date >= latest_date - months(6))

deaths_us_long <- deaths_us %>%
  pivot_longer(cols = -(1:11), names_to = "date", values_to = "deaths") %>%
  mutate(date = mdy(date)) %>%
  filter(date >= latest_date - months(6))

# Merge US cases and deaths data
covid_data_us <- left_join(cases_us_long, deaths_us_long, by = c("UID", "Province_State", "Country_Region", "date"))
```

## Summary Statistics

```{r summary-stats}
covid_summary_us <- covid_data_us %>%
  group_by(date) %>%
  summarise(
    total_cases = sum(cases, na.rm = TRUE),
    total_deaths = sum(deaths, na.rm = TRUE)
  )

# Display summary
head(covid_summary_us)
```

## State-wise Analysis (US Data)

```{r plot-states, fig.width=10, fig.height=6}
state_summary <- covid_data_us %>%
  filter(!is.na(Province_State)) %>%
  group_by(date, Province_State) %>%
  summarise(total_cases = sum(cases, na.rm = TRUE), .groups = "drop") %>%
  arrange(date)

state_summary <- state_summary %>% filter(!is.na(date))  # Ensure no NA values in date

# Reduce to top 10 states with the highest recent case counts
top_states <- state_summary %>%
  filter(date == max(date, na.rm = TRUE)) %>%
  arrange(desc(total_cases)) %>%
  slice_head(n = 10) %>%
  pull(Province_State)

state_summary_filtered <- state_summary %>% filter(Province_State %in% top_states)

plot_states <- ggplot(state_summary_filtered, aes(x = date, y = total_cases, color = Province_State)) +
  geom_line() +
  labs(title = "COVID-19 Cases by US States (Last 6 Months)", x = "Date", y = "Total Cases") +
  theme_minimal()

# Using static ggplot to reduce memory usage
plot_states
```
## State-wise Analysis (US Data_death_cases)

```{r plot-states, fig.width=10, fig.height=6}
state_summary <- covid_data_us %>%
  filter(!is.na(Province_State)) %>%
  group_by(date, Province_State) %>%
  summarise(total_deaths = sum(deaths, na.rm = TRUE), .groups = "drop") %>%
  arrange(date)

state_summary <- state_summary %>% filter(!is.na(date))  # Ensure no NA values in date

# Reduce to top 10 states with the highest recent case counts
top_states <- state_summary %>%
  filter(date == max(date, na.rm = TRUE)) %>%
  arrange(desc(total_deaths)) %>%
  slice_head(n = 10) %>%
  pull(Province_State)

state_summary_filtered <- state_summary %>% filter(Province_State %in% top_states)

plot_states <- ggplot(state_summary_filtered, aes(x = date, y = total_deaths, color = Province_State)) +
  geom_line() +
  labs(title = "COVID-19 Deaths by US States (Last 6 Months)", x = "Date", y = "Total Deaths") +
  theme_minimal()

# Using static ggplot to reduce memory usage
plot_states
```

## Conclusion

This report provides a reproducible analysis of COVID-19 trends across different US states using the Johns Hopkins University dataset. The analysis has been optimized for memory efficiency by limiting data to the last 6 months and focusing on the top 10 states.